### 2016-07-03 ansi-isolation.tex
Typos fixed.

Various isolation guarantees are defined as propositions indexed by
a transaction identifier $T_j$. Transaction $T_j$ is said to have
experienced \emph{read-my-writes} visibility under $\E$ if every
effect ($\eta_1$) of $T_j$ is visible to every subsequent effect
($\eta_2$) of the same transaction in $\E$. This lets $T_j$ to never
lose its own updates. Monotonic visibility adds one more constraint to
read-my-writes; besides requiring $\eta_1$ to be visible to $\eta_2$,
it also requires every effect ($\eta$) visible to $\eta_1$ in $\E$ to
be visible to $\eta_2$ as well. Thus $T_j$ witnesses monotonically
increasing state as time progresses. Atomic visibility allows an
effect $\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only
if all effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visiblity does not prevent an uncommitted
transaction from being visible. This is addressed by \C{CommitVis},
which requires the commit effect of $T_i$ to be visible whenever any
effect of $T_i$ is visible.

The ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
specification is therefore a combination of these two guarantees. The
specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
available replicated stores. On relational databases, however, {\sc
rc} has also come to be associated with the $\mathtt{MonotonicVis}$
guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
are sufficient to reason about {\sc rc} isolation on relational stores
too. The combination of these guarantees with the {\sc sc} property of
relational stores (formalized in \S\ref{sec:store-consistency})
automatically leads to the monotonicity guarantee, which explains why
{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
of the implementation. On weakly consistent stores however,
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply
$\mathtt{MonotonicVis}$. A stronger isolation level called
\iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores. The sample
{\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
{\sc mav}.

@@

\iso{Snapshot Isolation} ({\sc SI}) proscribes this possibility. If
$T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
insists that either $T_i$ be visible to $T_j$, or $T_j$ be visible to
the conflicting write of $T_i$ (this is captured by the auxiliary
definition \C{SnapshotSER}). Fig.~\ref{fig:ansi-iso-eg-si} shows an
execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
database that doesn't include $T_2$, but the conflicting write to $X$
in $T_2$, and the subsequent operations (due to the monotonicity
requirement of {\sc mav}) witness $T_1$. Thus, $T_2$ is aware of $T_1$
when it commits. Lastly, \iso{Serializable} isolation extends
$\mathtt{SnapshotSER}$ to also cover non-conflicting transactions that
write to variables read in the current transaction.
Fig.~\ref{fig:ansi-iso-eg-ser} shows an execution of an {\sc ser}
transacton $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
and $Y$, respectively. The snapshot witnessed by the {\sc ser}
transaction $T_3$ does not include $T_4$, but the write to $Y$ in
$T_4$, although non-conflicting, witnesses $T_1$ because $Y$ is read
by $T_1$. Visibility includes other operations of $T_4$ because of
{\sc mav}.

>>>>

Various isolation guarantees are defined as propositions indexed by
a transaction identifier $T_j$. Transaction $T_j$ is said to have
experienced \emph{read-my-writes} visibility under $\E$ if every
effect ($\eta_1$) of $T_j$ is visible to every subsequent effect
($\eta_2$) of the same transaction in $\E$. This lets $T_j$ to never
lose its own updates. Monotonic visibility adds one more constraint to
read-my-writes; besides requiring $\eta_1$ to be visible to $\eta_2$,
it also requires every effect ($\eta$) visible to $\eta_1$ in $\E$ to
be visible to $\eta_2$ as well. Thus $T_j$ witnesses monotonically
increasing state as time progresses. Atomic visibility allows an
effect $\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only
if all effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visibility does not prevent an uncommitted
transaction from being visible. This is addressed by \C{CommitVis},
which requires the commit effect of $T_i$ to be visible whenever any
effect of $T_i$ is visible.

The ANSI SQL 92 standard requires \iso{Read Committed} isolation to avoid
the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
specification is therefore a combination of these two guarantees. The
specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
available replicated stores. On relational databases, however, {\sc
rc} has also come to be associated with the $\mathtt{MonotonicVis}$
guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
are sufficient to reason about {\sc rc} isolation on relational stores
too. The combination of these guarantees with the {\sc sc} property of
relational stores (formalized in \S\ref{sec:store-consistency})
automatically leads to the monotonicity guarantee, which explains why
{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
of the implementation. On weakly consistent stores however,
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply
$\mathtt{MonotonicVis}$. A stronger isolation level called
\iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores. The sample
{\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
{\sc mav}.

@@

\iso{Snapshot Isolation} ({\sc SI}) proscribes this possibility. If
$T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
insists that either $T_i$ be visible to $T_j$, or $T_j$ be visible to
the conflicting write of $T_i$ (this is captured by the auxiliary
definition \C{SnapshotSER}). Fig.~\ref{fig:ansi-iso-eg-si} shows an
execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
database that doesn't include $T_2$, but the conflicting write to $X$
in $T_2$, and the subsequent operations (due to the monotonicity
requirement of {\sc mav}) witness $T_1$. Thus, $T_2$ is aware of $T_1$
when it commits. Lastly, \iso{Serializable} isolation extends
$\mathtt{SnapshotSER}$ to also cover non-conflicting transactions that
write to variables read in the current transaction.
Fig.~\ref{fig:ansi-iso-eg-ser} shows an execution of an {\sc ser}
transaction $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
and $Y$, respectively. The snapshot witnessed by the {\sc ser}
transaction $T_3$ does not include $T_4$, but the write to $Y$ in
$T_4$, although non-conflicting, witnesses $T_1$ because $Y$ is read
by $T_1$. Visibility includes other operations of $T_4$ because of
{\sc mav}.

### 2016-07-03 case-studies.tex
Typos fixed.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency, we
reimplemented the counter as a \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels, and ANSI SQL isolation levels. Keeping the
consistency level at {\sc ec}, we then repeated our experiments with
committed reads and serializable increments - a combination which was
proved to preserve monotonicity on an {\sc sc} store, but nonetheless
observed monotonicity violations. On the other hand, instantiating the
store consistency with {\sc cc} allowed us to obtain a proof, and
recover the monotonicity invariant. We in fact obtained a proof after
relaxing the serializability requirement for writes, by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
We implemented this version of the counter RDT, and obtained a 52\%
improvement in latency compared to the previous
implementation\footnote{The tradeoff is that the view count is now no
longer accurate, although it grows monotonically.}.

@@

Next, we focus our attention on a Twitter-like microblogging site that
is one the sample applications shipped with a popular textbook on
Ruby-on-Rails~\cite{railsbook}. One of the invariants that the
application chooses to enforce is the referential integrity between
microposts and authors. The application enforces this invariant
entirely in the user space by relying on database-backed transactions
(this behaviour is typical of Rails applications~\cite{bailisferal}).
For instance, deletion of a user is carried out in a transaction that
also includes the deletion of microposts authored by the user.
However, concurrent observers can nonetheless witness violations of
referential integrity. To measure the likelihood of such anomalies in
practice, we first populated a Postgres database with 1000 user
accounts, each user with 50 microposts and 20 followers, choosen
uniformly at random. We then constructed experiments with 64
concurrent clients performing deletions and timeline reads of randomly
choosen users in 1:7 ratio, and witnessed a maximum of 250 violations
of referential integrity. We then instantiated our framework for an
{\sc sc} store with various ANSI SQL isolation levels for user
deletions and timeline reads, and obtained a proof that {\sc rc} is
sufficient for deletions, whereas reads require {\sc rr}. The proof is
immediately applicable for MySQL, which implements ANSI SQL isolation
levels. Since Postgres's isolation levels are stronger than the
corresponding ANSI levels, \rulelabel{RG-Conseq} rule
(Fig.~\ref{fig:rg-rules}) lets us deduce that the same choice of
isolation levels also work on Postgres. Indeed, experiments confirmed
this judgment.

>>>>

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency, we
re-implemented the counter as a \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels, and ANSI SQL isolation levels. Keeping the
consistency level at {\sc ec}, we then repeated our experiments with
committed reads and serializable increments - a combination which was
proved to preserve monotonicity on an {\sc sc} store, but nonetheless
observed monotonicity violations. On the other hand, instantiating the
store consistency with {\sc cc} allowed us to obtain a proof, and
recover the monotonicity invariant. We in fact obtained a proof after
relaxing the serializability requirement for writes, by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
We implemented this version of the counter RDT, and obtained a 52\%
improvement in latency compared to the previous
implementation\footnote{The trade-off is that the view count is now no
longer accurate, although it grows monotonically.}.

@@

Next, we focus our attention on a Twitter-like microblogging site that
is one the sample applications shipped with a popular textbook on
Ruby-on-Rails~\cite{railsbook}. One of the invariants that the
application chooses to enforce is the referential integrity between
microposts and authors. The application enforces this invariant
entirely in the user space by relying on database-backed transactions
(this behaviour is typical of Rails applications~\cite{bailisferal}).
For instance, deletion of a user is carried out in a transaction that
also includes the deletion of microposts authored by the user.
However, concurrent observers can nonetheless witness violations of
referential integrity. To measure the likelihood of such anomalies in
practice, we first populated a Postgres database with 1000 user
accounts, each user with 50 microposts and 20 followers, chosen
uniformly at random. We then constructed experiments with 64
concurrent clients performing deletions and timeline reads of randomly
chosen users in 1:7 ratio, and witnessed a maximum of 250 violations
of referential integrity. We then instantiated our framework for an
{\sc sc} store with various ANSI SQL isolation levels for user
deletions and timeline reads, and obtained a proof that {\sc rc} is
sufficient for deletions, whereas reads require {\sc rr}. The proof is
immediately applicable for MySQL, which implements ANSI SQL isolation
levels. Since Postgres's isolation levels are stronger than the
corresponding ANSI levels, \rulelabel{RG-Conseq} rule
(Fig.~\ref{fig:rg-rules}) lets us deduce that the same choice of
isolation levels also work on Postgres. Indeed, experiments confirmed
this judgment.

### 2016-07-03 introduction.tex
Typos fixed.

In this paper, we propose a program logic for weakly-isolated
transactions that realizes this goal.  In particular, we develop a set
of syntax-directed compositional proof rules that allow developers to
build correctness proofs for transactional programs in the presence of
a weakly isolated concurrency control mechanism.  A key novelty of our
approach is that it is parametric over the \emph{isolation semantics}
of transactions in the program, as well as the \emph{consistency
  semantics} of the underlying store. In concrete terms, this means
that, unlike recent work focussed on formalizing the semantics of weak
consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
does not assume a minimum or predefined set of consistency or
isolation levels. Instead, our operational semantics admits
declarative specifications of transaction isolation and store
consistency levels (\eg~\cite{pldi15,gotsmanconcur15}), and generates
executions that are guaranteed to satisfy these specifications. The
result is a flexible system model that is general enough to
incorporate the semantics of a range of isolation levels on a variety
of stores (\eg the sequentially consistent store of~\cite{adyaphd}, or
the causally consistent store of~\cite{gotsmanpopl16}, etc).  Our key
technical contribution is thus a set of proof rules that demonstrate
that a program with a given selection of isolation levels for its
transactions preserves its invariants when executed on a store
equipped with certain consistency guarantees (e.g., causal
visibility).

The paper makes the following contributions:
\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly-isolated transactions, demonstrating that a general
    parameteric isolation semantics can be expressed as
    well-formedness constraints on a program's execution.
  \item We propose a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are associated with specific weak-isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

>>>>

In this paper, we propose a program logic for weakly-isolated
transactions that realizes this goal.  In particular, we develop a set
of syntax-directed compositional proof rules that allow developers to
build correctness proofs for transactional programs in the presence of
a weakly isolated concurrency control mechanism.  A key novelty of our
approach is that it is parametric over the \emph{isolation semantics}
of transactions in the program, as well as the \emph{consistency
  semantics} of the underlying store. In concrete terms, this means
that, unlike recent work focused on formalizing the semantics of weak
consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
does not assume a minimum or predefined set of consistency or
isolation levels. Instead, our operational semantics admits
declarative specifications of transaction isolation and store
consistency levels (\eg~\cite{pldi15,gotsmanconcur15}), and generates
executions that are guaranteed to satisfy these specifications. The
result is a flexible system model that is general enough to
incorporate the semantics of a range of isolation levels on a variety
of stores (\eg the sequentially consistent store of~\cite{adyaphd}, or
the causally consistent store of~\cite{gotsmanpopl16}, etc).  Our key
technical contribution is thus a set of proof rules that demonstrate
that a program with a given selection of isolation levels for its
transactions preserves its invariants when executed on a store
equipped with certain consistency guarantees (e.g., causal
visibility).

The paper makes the following contributions:
\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly-isolated transactions, demonstrating that a general
    parametric isolation semantics can be expressed as
    well-formedness constraints on a program's execution.
  \item We propose a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are associated with specific weak-isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

### 2016-07-03 opsem.tex
edits to sec. 5
Typos fixed.

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace
($\E$). The first component of the trace is a set ($\A$) of
\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
performed during the execution. Any value associated with the
operation (\eg, value read or value written) is also documented, and
can be accessed via $\rval$. Every effect has a unique identifier
accessible via $\id$, and a transction identifier accessible via
$\txn$.  The latter identifies the transaction that generated the
effect. In every step of the evaluation, the machine reduces a \txnimp\
term by executing a read, write or commit operation, generating an
effect, and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

@@

A more notable source of non-determinism, however, is the
\rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly-consistent data stores. For instance, operations issued to an
eventually consistent ({\sc ec}) replicated store could be dispatched
to different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores; we elaborate on the implication of this style of
definition in \S\ref{sec:store-consistency}.  Stronger visiblity
properties can be expressed by imposing well-formedness constraints
over $\visZ$ via the trace invariant ($\I$). Since the abstract
machine is obligated to satisfy $\I$ at every step of the execution,
operations are guaranteed to experience the level of isolation
specified by $\I$.  Thus, in executions that run to completion, the
abstract machine models a store that provides the required levels of
isolation.  Notably, the machine achieves this without defining an
operationalsemantics for isolation levels, instead solely relying on
their declarative characterization as trace well-formedness
constraints to enforce isolation guarantees.
\S\ref{sec:ansi-isolation} shows the specification of various ANSI SQL
isolation levels stated as trace well-formedness constraints.

>>>>

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace
($\E$). The first component of the trace is a set ($\A$) of
\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
performed during the execution. Any value associated with the
operation (\eg, value read or value written) is also documented, and
can be accessed via $\rval$. Every effect has a unique identifier
accessible via $\id$, and a transaction identifier accessible via
$\txn$.  The latter identifies the transaction that generated the
effect. In every step of the evaluation, the machine reduces a \txnimp\
term by executing a read, write or commit operation, generating an
effect, and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

@@

A more notable source of non-determinism, however, is the
\rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly-consistent data stores. For instance, operations issued to an
eventually consistent ({\sc ec}) replicated store could be dispatched
to different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores; we elaborate on the implication of this style of
definition in \S\ref{sec:store-consistency}.  Stronger visibility
properties can be expressed by imposing well-formedness constraints
over $\visZ$ via the trace invariant ($\I$). Since the abstract
machine is obligated to satisfy $\I$ at every step of the execution,
operations are guaranteed to experience the level of isolation
specified by $\I$.  Thus, in executions that run to completion, the
abstract machine models a store that provides the required levels of
isolation.  Notably, the machine achieves this without defining an
operational semantics for isolation levels, instead solely relying on
their declarative characterization as trace well-formedness
constraints to enforce isolation guarantees.
\S\ref{sec:ansi-isolation} shows the specification of various ANSI SQL
isolation levels stated as trace well-formedness constraints.
